---
title: "Prevention is Better than Cure—Statistical Observations of Cardiovascular Disease Data"
author: "Ty Kentaro Okazaki, Matthew R Herradura, David Xie, Carson Klein"
date: "`r Sys.Date()`"
output: openintro::lab_report
---

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(infer)
```

```{r load-data, message=FALSE}
cms_data <- read.csv("../data/cms_data.csv")
```

## Introduction

Using data gathered from Medicare and Medicaid patients, the Center for Medicare & Medicaid services (CMS) compiles claim data across numerous categorical and numerical variables. Collected from a variety of sources including: hospital databases, diagnostic reports, and surveys whilst adhering to an array of Quality Rating System (QRS) guidelines and requirements, the Center for Medicare & Medicaid services provides reliable and representative data. Computed by personnel in CDC's Division for Heart Disease and Stroke Prevention (DHDSP), furthermore, this [data set](https://chronicdata.cdc.gov/Heart-Disease-Stroke-Prevention/Center-for-Medicare-Medicaid-Services-CMS-Medicare/iw6q-r3ja) provides potential insights on Cardiovascular disease (CVD) indicators and associated risk factors; categorized by variables such as age, race/ethnicity, and location, this data set can be processed and analyzed using statistical methods to highlight insights into CVD prevention. Specifically, this data set includes anonymous CVD hospitalizations data from Medicare & Medicaid patients of age 65+ across the United States. Considering the influences and patterns variables can have with regard to predicting CVD hospitalization instances, it is important to dissect and analyze those of paramount importance. Consequently, our general question(s) are as follows: What are the main factors that influence a patient's susceptibility for developing cardiovascular disease? Is there a significant relationship between specific variables and the prevalence of cardiovascular disease?

Null Hypothesis: Predictor variables such as age, gender, and ethnicity will have no influence on a patient's susceptibility for developing cardiovascular disease.

Alternative Hypothesis: Predictor variables such as age, gender, and ethnicity will influence a patient's susceptibility for developing cardiovascular disease.

## Analysis

#### How is race related to Cardiovascular Disease and Hospitalizations?

#### 

#### How does geographic location influence Cardiovascular Disease and Hospitalizations?

#### 

## Conclusion
